Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2024-01-04 demystify.tex
updates

This document is available at \url{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf} and its \LaTeX{} source is on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{Github}. If you have questions or comments, feel free to create a \href{https://github.com/nguyenthanhvuh/phd-cs-us/issues}{GitHub issue} for discussion.

@@

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories on how students are mistreated, stressed, and miserable. Certainly issues including bad relationships with professors, conflicts with co-authors and lab mates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student) \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, happy students, etc.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

This section lists some cultural issues that international students might want to pay attention to. These issues are not specific to just the PhD admission process but in general.

@@

Students, especially international ones, often get confused on what to do after they get an offer from a professor, e.g., how to postpone or accept/reject offers, and most importantly, how to do so without offending anyone. In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept it. If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This is considered rude and unprofessional.  If you do so, you will likely burn bridges with that professor.

@@

This is very similar to a professional job offer.  You should be serious about it and communicate promptly. And most importantly, do not accept an offer and then later reject it.  If you do this with a company, you might face legal consequences.  If you do this with a professor, they will likely not do anything about it (i.e., not worth their time) but they will not want to work with you in the future.

@@

\section{Illegal Software} Using illegal/cracked software is very common in many countries (and even in the US). However, \emph{do not} install or use them on university computers (e.g., those given by your department or adviser).  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or adviser being sued for a large sum of money, and it is \emph{you} who caused it.  If you need software, ask your adviser or the department.

\section{Gifts} In many countries, it is customary to give professors gifts, often during holidays.  These gifts can be costly and profs. sometimes expected them. In the US, it is uncommon and perfectly OK to not to give gifts. \emph{However}, if you'd like to offer small souvenir-like tokens, it's a thoughtful gesture that's appreciated. Some professors proudly display their gifts, which can come from students and colleagues (e.g., when they travel to their home countries or conferences). In summary, small gifts are fine, but avoid anything that might make your professors uncomfortable.

\section{Relationships with your Profs.}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends.% These connections are genuine and meaningful.

>>>>

This document is available at

\begin{center}
\url{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf}, 
\end{center}

\noindent and its \LaTeX{} source is also on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{Github}. If you have questions or comments, feel free to create a \href{https://github.com/nguyenthanhvuh/phd-cs-us/issues}{GitHub issue} for discussion.

@@

\section{Will I be miserable during my a PhD?}\label{sec:happy}
There are many stories on how students are mistreated, stressed, and miserable. Issues including bad relationships with professors, conflicts with co-authors and lab mates, feeling discriminated (e.g., because you're an international student), etc \emph{do} exist, and it is good to be aware of those.  However, in reality there are many good mentors, supportive lab mates and department, and so on.  So don't let social media make you feel pessimistic and deter your quest to advance knowledge.

@@

This section lists some cultural issues that international students might want to pay attention to. These issues might not specific to just the PhD admission process but in general.

@@

Students, especially international ones, often get confused on what to do after they get an offer from a professor, e.g., how to postpone or accept/reject offers, and most importantly, how to do so without offending anyone. In short, if you are not serious about an offer for any reason (e.g., still waiting for other offers), do not accept, i.e., commit to it. If you are not sure, ask for more time to decide.  But do not accept an offer and then later reject it.  This is considered \emph{rude and unprofessional}, and will likely burn bridges with that professor.

@@

This is very similar to a professional job offer.  You should be serious about it and communicate promptly. Importantly, do not accept an offer and then later reject it.  If you do this with a company, you might face legal consequences.  If you do this with a professor, they will likely not do anything about it (i.e., not worth their time) but they will not want to work with you in the future.

@@

\section{Illegal Software} Using illegal/cracked software is very common in many countries (and even in the US). However, \emph{do not} install or use them on university computers, even those given to you from your adviser.  It is unlikely that the university will track you down, but it is the \emph{software company} that will.  They have very sophisticated tools to detect illegal software and will sue your university/department.  Imagine your department or adviser being sued for a large sum of money, and it is \emph{you} who caused it.  If you need to purchase software,  ask your adviser or the department.

\section{Gifts}\label{sec:gifts} In many countries, it is customary to give professors gifts, often during holidays.  These gifts can be costly and profs. sometimes expected them. In the US, it is uncommon and perfectly OK to not to give gifts. \emph{However}, if you'd like to offer small souvenir-like tokens, it's a thoughtful gesture that's appreciated. Some professors proudly display their gifts, which can come from students and colleagues (e.g., when they travel to their home countries or conferences). In summary, small gifts are fine, but avoid anything that might make your professors uncomfortable.

\section{Relationships with your Profs.}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks (\S\ref{sec:gifts}).  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends. %These connections are genuine and meaningful.
